---
title: "Project_Data_Preparation"
author: "Megan Glass"
date: "2023-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries 
```{r}
library(tidyverse)
#install.packages("plotmo")
library(plotmo)
#install.packages("readxl")
library(readxl)

```

##Load the data 
```{r}
jump_data <-read_excel("Jump_Data.xlsx")
```

##Cleaning data

Taking out columns that are not necessary 
```{r}
sum(is.na(jump_data))
#509 - column for body weight

jump_data <- jump_data[,-c(2,6)]
jump_data

sum(is.na(jump_data))
#0 

```


Making every athlete its own data set - to look for individual differences and looking at summary statistics for each 
```{r}
athlete1 <- jump_data[c(19:51), ]
summary(athlete1$`Jump Height (Flight Time) in Inches [in]`)

athlete2 <- jump_data[c(70:102), ]
summary(athlete2$`Jump Height (Flight Time) in Inches [in]`)

athlete3 <- jump_data[c(120:152), ]
summary(athlete3$`Jump Height (Flight Time) in Inches [in]`)

athlete4 <- jump_data[c(168:197), ]
summary(athlete4$`Jump Height (Flight Time) in Inches [in]`)

athlete5 <- jump_data[c(225:251), ]
summary(athlete5$`Jump Height (Flight Time) in Inches [in]`)

athlete6 <- jump_data[c(270:302), ]
summary(athlete6$`Jump Height (Flight Time) in Inches [in]`)

#athlete 7 has one jump date we excluded since the weight was drastically different 
athlete7 <- jump_data[c(321:338, 342:347), ]
summary(athlete7$`Jump Height (Flight Time) in Inches [in]`)

athlete8 <- jump_data[c(369:398), ]
summary(athlete8$`Jump Height (Flight Time) in Inches [in]`)

athlete9 <- jump_data[c(429:461), ]
summary(athlete9$`Jump Height (Flight Time) in Inches [in]`)

athlete10 <- jump_data[c(480:509), ]
summary(athlete10$`Jump Height (Flight Time) in Inches [in]`)
```
Combining all athlete's data into one new data set
```{r}
new_jump_data <- rbind(athlete1, athlete2, athlete3, athlete4, athlete5, athlete6, athlete7, athlete8, athlete9, athlete10)
#names(new_jump_data) <- make.names(names(jump_data))
```

Reverse the order of the data frame to have it go from Janruary to April
```{r}
new_jump_data <- new_jump_data[nrow(new_jump_data):1,]
```

Mean difference from previous week for jump height and flight time:contraction time
```{r}
aths <- unique(new_jump_data[,c("Athlete", "Test Date")])

mean_diff_flight_time <- rep(NA, nrow(new_jump_data))
mean_diff_jump_height <- rep(NA, nrow(new_jump_data))
mean_diff_concentric_impulse <- rep(NA, nrow(new_jump_data))
mean_diff_eccentric_duration <- rep(NA, nrow(new_jump_data))
mean_diff_force_zero <- rep(NA, nrow(new_jump_data))
mean_diff_assymetry <- rep(NA, nrow(new_jump_data))
mean_diff_peak_power <- rep(NA, nrow(new_jump_data))

for(i in 2:nrow(new_jump_data)){
  if(new_jump_data$Athlete[i] == new_jump_data$Athlete[i-1] &
     new_jump_data$`Test Date`[i] != new_jump_data$`Test Date`[i-1]){
    
    mean_diff_flight_time[i:(i+2)] <- new_jump_data$`Flight Time:Contraction Time`[i:(i+2)]/mean(new_jump_data$`Flight Time:Contraction Time`[(i-3):(i-1)], na.rm = TRUE)
    
     mean_diff_jump_height[i:(i+2)] <- new_jump_data$`Jump Height (Flight Time) in Inches [in]`[i:(i+2)]/mean(new_jump_data$`Jump Height (Flight Time) in Inches [in]`[(i-3):(i-1)], na.rm = TRUE)
     
     mean_diff_concentric_impulse[i:(i+2)] <- new_jump_data$`Concentric Impulse [N s]`[i:(i+2)]/mean(new_jump_data$`Concentric Impulse [N s]`[(i-3):(i-1)], na.rm = TRUE)
     
      mean_diff_eccentric_duration[i:(i+2)] <- new_jump_data$`Eccentric Duration [ms]`[i:(i+2)]/mean(new_jump_data$`Eccentric Duration [ms]`[(i-3):(i-1)], na.rm = TRUE)
      
      mean_diff_force_zero[i:(i+2)] <- new_jump_data$`Force at Zero Velocity [N]`[i:(i+2)]/mean(new_jump_data$`Force at Zero Velocity [N]`[(i-3):(i-1)], na.rm = TRUE)
      
      mean_diff_assymetry[i:(i+2)] <- new_jump_data$`Eccentric Deceleration RFD Asymmetry [% L,R]`[i:(i+2)]/mean(new_jump_data$`Eccentric Deceleration RFD Asymmetry [% L,R]`[(i-3):(i-1)], na.rm = TRUE)
      
      mean_diff_peak_power[i:(i+2)] <- new_jump_data$`Eccentric Peak Power / BM [W/kg]`[i:(i+2)]/mean(new_jump_data$`Eccentric Peak Power / BM [W/kg]`[(i-3):(i-1)], na.rm = TRUE)

    
  }
}

new_jump_data$mean_diff_flight_time <- mean_diff_flight_time
new_jump_data$mean_diff_jump_height <- mean_diff_jump_height
new_jump_data$mean_diff_concentric_impulse <- mean_diff_concentric_impulse
new_jump_data$mean_diff_eccentric_duration <- mean_diff_eccentric_duration
new_jump_data$mean_diff_force_zero <- mean_diff_force_zero
new_jump_data$mean_diff_assymetry<- mean_diff_assymetry
new_jump_data$mean_diff_peak_power<- mean_diff_peak_power
```

Mean of next weeks jumps
```{r}
#aths <- unique(new_jump_data[,c("Athlete", "Test Date")])

next_weeks_jump_height <- rep(NA, nrow(new_jump_data))

for(i in 2:(nrow(new_jump_data)-2)){
  if(new_jump_data$Athlete[i] == new_jump_data$Athlete[i+1] &
     new_jump_data$`Test Date`[i] != new_jump_data$`Test Date`[i+1]){
    
      next_weeks_jump_height[i:(i-2)] <-mean(new_jump_data$`Jump Height (Flight Time) in Inches [in]`[(i+1):(i+3)], na.rm = TRUE)
    
  }
}

new_jump_data$next_weeks_jump_height <- next_weeks_jump_height
```


Deviations for Jump Height and Contraction Time from overall average
```{r}
jump_height_deviations <- rep(NA, nrow(new_jump_data))
flight_time_deviations <- rep(NA, nrow(new_jump_data))


for (i in unique(new_jump_data$Athlete)){
  jump_height_deviation <- new_jump_data[which(new_jump_data$Athlete==i),]$`Jump Height (Flight Time) in Inches [in]` - mean(new_jump_data[which(new_jump_data$Athlete==i),]$`Jump Height (Flight Time) in Inches [in]`)
  
  jump_height_deviations[which(new_jump_data$Athlete==i)]<- jump_height_deviation
  
  flight_time_deviation <- new_jump_data[which(new_jump_data$Athlete==i),]$`Flight Time:Contraction Time` - mean(new_jump_data[which(new_jump_data$Athlete==i),]$`Flight Time:Contraction Time`)
  
  flight_time_deviations[which(new_jump_data$Athlete==i)] <- flight_time_deviation
}

new_jump_data$jump_height_deviations <- jump_height_deviations
new_jump_data$flight_time_deviations <- flight_time_deviations
```


